\textsc{Regression And Resampling Techniques \\ FYS-STK4155 \\ Assignment One}

### 2019-09-10 assignment_one/article/main.tex
Working on article.

\documentclass[article, a4paper, oneside, 12pt]{memoir}
\linespread{1.25}
%\setlength{\parskip}{1em}
%\setlength{\parindent}{0em}

\usepackage[]{mathtools}
\usepackage[]{amssymb}
\usepackage[]{libertine}
\usepackage[]{libertinust1math}
\usepackage[T1]{fontenc}
\usepackage[]{microtype}
\usepackage[]{geometry}
\usepackage[]{amsthm}
\usepackage[]{thmtools}
\usepackage[usedvipsnames]{xcolor}
\usepackage[]{bm}
\usepackage[]{amsmath}
\usepackage[]{commath}
\usepackage[noabbrev, capitalize]{cleveref}

\declaretheorem[style=remark, name=Remark,
mdframed={
}]{remark}

@@

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}
\newcommand{\diag}[1]{\mathrm{diag}(#1)}
\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
\newcommand{\R}[1]{\mathrm{R}^2(#1)}

\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\y}{\bm{y}}
\newcommand{\data}{\bm{\Omega}}
\newcommand{\cost}{\mathcal{C}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\basis}{\mathcal{B}}
\newcommand{\set}[1]{\left\{\, #1 \right\}}

@@

@@

  Regression analysis is the act of estimating relationships among variables.
  In this project, we study various regression methods in more detail. In
  particular, we compare the \emph{ordinary least squares} (OLS) method with
  the \emph{Ridge regression} and \emph{Lasso regression} techniques. As we
  shall see, these three methods are all variations over the same theme. We
  start by testing the methods on noisy data sampled from a function known as
  \emph{Franke's} bivariate test function, which has been in widespread use in
  the evaluation of interpolation and data fitting techniques. Finally, we run
  regression on real terrain data, comparing the aforementioned methods.

@@

  In general, the goal of regression analysis is to fit a \emph{model function}
  \( f(\x, \beta) \) to a set of \( n \) data points \( \data = (\x_i,
  y_i)_{i=1}^n \). A simple example is that of a linear polynomial with two
  parameters:
  \begin{equation}
    f(x, \beta) = \beta_0 + \beta_1 x.
  \end{equation}
  The \emph{model parameters} \( \beta \) are determined in order to minimize a
  suitable \emph{cost function} \(\cost(\data, \beta)\) which measures to which
  extent the model function manages to capture trends in the data \( \data \).
  It is the choice of cost function \( \matcal{C}(\data, \beta)\) which
  distinguishes  the three regression techniques, OLS, Lasso regression, and
  Ridge regression.

  \begin{remark}
    It is often assumed a priori that the data is infact generated from a noisy
    model, such that each \( y_i \) can be described as
    \begin{equation}
      y_i = f(\x_i) + \varepsilon_i
    \end{equation}
    where each \( \varepsilon_i \sim \N(0, \sigma^2) \) is normally distributed
    with zero mean and variance \( \sigma^2 \). This assumption on the error
    gives rise to what is known as \emph{general linear models}.
  \end{remark}

@@

  \paragraph{Optimizing the parameters in OLS.}

@@

  We have that
  \begin{align}
    \dpd{\cost(\data, \beta)}{\beta} = \X^T(\y - \X\beta), 
  \end{align}
  and equating this to zero yields the familiar \emph{normal equations}
  \begin{equation}
    \X^T \y = \X^T\X\beta.
  \end{equation}
  If the matrix \( \X^T\X \) is invertible, we may obtain the solution by
  direct numerical matrix inversion. In this case, the optimal model parameters
  are found directly by
  \begin{equation}
    \label{eq:optimal}
    \beta = (\X^T\X)^{-1}\X^T\y.
  \end{equation}
  However, these matrices may be ill-conditioned when the number of equations
  are very large, and it is therefore common to apply approximate solvers for
  the inverse, for instance using low-rank approximation based on
  \textsc{SVD}-decomposition, which we will briefly turn to in the following.

  \paragraph{Singular value decomposition.}

@@

>>>>

\documentclass[dvipsnames, article, a4paper, oneside, 12pt]{memoir}
\linespread{1.25}
%\setlength{\parskip}{1em}
%\setlength{\parindent}{0em}

\usepackage[]{mathtools}
\usepackage[]{amssymb}
\usepackage[]{libertine}
\usepackage[]{libertinust1math}
\usepackage[T1]{fontenc}
\usepackage[]{microtype}
\usepackage[]{geometry}
\usepackage[]{amsthm}
\usepackage[]{thmtools}
\usepackage[usedvipsnames]{xcolor}
\usepackage[]{bm}
\usepackage[]{amsmath}
\usepackage[]{commath}
\usepackage[]{todonotes}
\usepackage[noabbrev, capitalize]{cleveref}
\usepackage[]{graphicx}
\usepackage{titling}

\declaretheorem[style=remark, name=Remark,
mdframed={
}]{remark}
\declaretheorem[style=remark, name=Assumption,
mdframed={
}]{assumption}

@@

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}
\newcommand{\diag}[1]{\mathrm{diag}(#1)}
\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
\newcommand{\R}[1]{\mathrm{R}^2(#1)}
\newcommand{\expect}{\mathbb{E}}
\newcommand{\OLS}{\mathrm{OLS}}
\newcommand{\ridge}{\mathrm{ridge}}

\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\y}{\bm{y}}
\newcommand{\data}{\bm{\Omega}}
\newcommand{\cost}{\mathcal{C}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\basis}{\mathcal{B}}
%\newcommand{\set}[1]{\left\{\, #1 \right\}}

\pretitle{%
  \begin{center}
    \includegraphics[scale=0.7]{images/franke_shadow.pdf}\\[\bigskipamount]
}
\posttitle{\end{center}}

@@

  \tableofcontents

@@

  Regression analysis is the act of estimating relationships among variables.
  In this project, we study various regression methods in more detail. In
  particular, we compare the \emph{ordinary least squares} (OLS) method with
  the \emph{Ridge regression} and \emph{Lasso regression} techniques. As we
  shall see, these three methods are all variations over the same theme. We
  start by testing the methods on noisy data sampled from a function known as
  \emph{Franke's} bivariate test function (visualized on the title page), which
  has been in widespread use in the evaluation of interpolation and data
  fitting techniques. Finally, we run regression on real terrain data,
  comparing the aforementioned methods.

@@

  In general, the goal of regression analysis is to fit a \emph{model function}
  \( f(\x, \beta) \) to a set of \( n \) data points \( \data = (\x_i,
  y_i)_{i=1}^n \). A simple example is that of a linear polynomial with two
  parameters:
  \begin{equation}
    f(x, \beta) = \beta_0 + \beta_1 x.
  \end{equation}
  The \emph{model parameters} \( \beta \) are determined in order to minimize a
  suitable \emph{cost function} \(\cost(\data, \beta)\) which measures to which
  extent the model function manages to capture trends in the data \( \data \).
  It is the choice of cost function \( \mathcal{C}(\data, \beta)\) which
  distinguishes  the three regression techniques, OLS, Lasso regression, and
  Ridge regression.

  \begin{assumption}
    \label{as:general_linear}
    It is often assumed a priori that the data is infact generated from a noisy
    model, such that each \( y_i \) can be described as
    \begin{equation}
      y_i = f(\x_i) + \varepsilon_i
    \end{equation}
    where each \( \varepsilon_i \sim \N(0, \sigma^2) \) is normally distributed
    with zero mean and variance \( \sigma^2 \). This assumption on the error
    gives rise to what is known as \emph{general linear models}.
  \end{assumption}

@@

  \subsection{Optimizing the parameters in OLS.}

@@

  We have that
  \begin{align}
    \dpd{\cost(\data, \beta)}{\beta} = \X^T(\y - \X\beta), 
  \end{align}
  and equating this to zero yields the familiar \emph{normal equations}
  \begin{equation}
    \X^T \y = \X^T\X\beta.
  \end{equation}
  If the matrix \( \X^T\X \) is invertible, we may obtain the solution by
  direct numerical matrix inversion. In this case, the optimal model parameters
  are found directly by
  \begin{equation}
    \label{eq:optimal}
    \beta_{\OLS} = (\X^T\X)^{-1}\X^T\y.
  \end{equation}
  However, these matrices may be ill-conditioned when the number of equations
  are very large, and it is therefore common to apply approximate solvers for
  the inverse, for instance using low-rank approximation based on
  \textsc{SVD}-decomposition, which we will briefly turn to in the following.

  \subsection{Singular value decomposition.}

@@

  \subsection{The bias-variance tradeoff}

  Since we are attempting to estimate a true distribution based on finite
  samples, there are certain statistical properties it is worthwile discussing.
  We follow the derivation in.  Recall \cref{as:general_linear} which states
  that our true data is generated from a noisy model
  \begin{equation}
    y = f(\x) + \varepsilon
  \end{equation}
  with \( \varepsilon \sim \N(0, \sigma^2) \).  We select a finite sample from
  the true data, which constitutes our finite data-set \( \data \).
  Furthermore, we have a model function \( f(\data, \beta) \) which depends on
  the selected data-set and the model parameters \( \beta \). Thus, our
  cost-function \( \cost \) will depend on the specific sample taken from the
  true distribution. It is of interest to examine the expected out-of-sample
  error \( \expect_{\data, \varepsilon}[\cost(\data, \beta)] \) of the model
  function over the sampled data set \( \data \) and the noise \( \varepsilon
  \). This can be decomposed into three distinct parts,
  \emph{\textcolor{Maroon}{variance}}, \emph{\textcolor{Blue}{bias}}, and
  \emph{\textcolor{Green}{noise}}:
  \begin{equation}
    \expect_{\data, \varepsilon}[\cost(\data, \beta)] = \frac{1}{n} \textcolor{Blue}{\sum_{i=1}^n \left(f(\x_i) - \expect_{\data}\left[\hat{y}_i\right]\right)^2}  + \frac{1}{n}\textcolor{Maroon}{\sum_{i=1}^n \expect_{\data}\big[\left(\hat{y}_i - \expect_{\data} [\hat{y}_i]\right)^2\big]} + \textcolor{Green} \sigma^2.
  \end{equation}
  See \cref{ap:derivation} for the full derivation. The bias-term measures the
  deviation of the expectation of the model function \(f(\data, \beta)\) from
  the true value \( f \), while the variance-term measures the fluctuations
  arising due to using finite samples.

  \section{Ridge Regression}

  When the dimensionality of the matrix \( \X \) is high, the problem of
  near-collinear columns often arise, as mentioned earlier. Thus, the model
  parameters \( \beta_{\OLS} \) of the ordinary least squares regression cannot
  be determined. The propsed fix to this problem is to slightly perturb the
  matrix \( \X^T \X \) by adding small positive values along the diagonal to
  ensure its invertability. This is known as \emph{ridge regression} and yields
  a model parameter estimator
  \begin{equation}
    \beta_{\ridge}(\lambda) = (\X^T\X + \lambda\mat{I})^{-1}\X^T\y
  \end{equation}
  where \( \lambda \geq 0 \) is a hyper parameter. To see why the resulting
  modified matrix is invertible, one has to look at the singular values.  It
  turns out that the model parameters \( \beta_{\ridge}(\lambda) \) directly
  minimize the \emph{ridge loss function} defined as
  \begin{equation}
    \cost_{\ridge}(\data, \beta;  \lambda) \coloneqq \frac{1}{n} \sum_{i=1}^n (y_i - \hat{y}_i)^2 + \lambda \sum_{i=1}^p \beta_i^2 = \cost(\data, \beta) + \lambda \beta^T \beta,
  \end{equation}
  which we recognize as standard penalized least squares.

  Ridge regregression carries several benefits. One of the more important is
  its resilience to overfitting. In a high-dimensional setting, if many of the
  data-features can be attributed to a few covariates, then the remaining
  covariates do nothing more than fit to the noise. Large values of \( \beta_i
  \) then a good indication that your model might be overfitting as some
  covariates are given more weight than others.  The Ridge regression solves
  this by directly penalizing large model parameters in the cost function.

  \appendix
  \chapter{Derivation of the bias-variance-noise relationship}
  \label{ap:derivation}

  \if 0
  To this end --- by adding and subtracting \( f(\x_i) \) --- we have that 
  \begin{align}
    \begin{split}
      \expect_{\data, \varepsilon} \big[\cost(\data, \beta)\big] &= \frac{1}{n}\expect_{\data, \varepsilon} \left[\sum_{i = 1}^n (y_i - \hat{y}_i)^2\right] \\
                                                                 &= \frac{1}{n}\expect_{\data, \varepsilon} \left[\sum_{i=1}^n (y_i - f(\x_i) + f(\x_i) - \hat{y}_i)^2\right] \\
                                                                 &= \frac{1}{n}\sum_{i=1}^n\left\{\overbrace{\expect_{\varepsilon} \left[(y_i - f(\x_i))^2\right]}^{= \sigma^2} + \expect_{\data}\left[(f(\x_i) - \hat{y}_i)^2\right] + 2\overbrace{\expect_{\varepsilon}\left[y_i - f(\x_i)\right]}^{= 0}\expect_{\data}\left[f(\x_i) - \hat{y}_i\right] \right\} \\ 
                                                                 &= \frac{1}{n}\left\{n\sigma^2 + \expect_{\data} \left[(f(\x_i) - \hat{y}_i)^2\right]\right\} \\
                                                                 &= \sigma^2 + \frac{1}{n}\sum_{i=1}^n\expect_{\data} \left[(f(\x_i) - \hat{y}_i)^2\right]\right\}.
  \end{split}
  \end{align}

  Using the same trick of adding and subtracting, we may further decompose the second term as
  \begin{align}
    \begin{split}
      \expect_{\data} \big[ (f(\x_i) - \hat{y}_i)^2\big] = \left(f(\x_i) - \expect_{\data}\big[\hat{y_i}\big]\right)^2 + \expect_{\data} \left[\left(\hat{y}_i - \expect_{\data}\left[\hat{y}_i\right]\right)^2\right]
    \end{split}
  \end{align}
  \fi
  Combining the two previous equations, we obtain
